---
title: "Graphically Pawsitive Vibes"
subtitle: "Proposal"
author: 
  - name: "Isle of Dogs"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project will graphically depict trends, patterns, and analysis of Animal Shelter data derived from the TidyTuesday initiative. The dataset includes information about animals taken in and released from the shelter; there are variables such as intake type, outcome type, animal type, breed, age, and more. The primary goal of this project is to identify key factors that influence an animals outcome."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
#| echo: false

if (!require("pacman")) 
  install.packages("pacman, formattable, diagnose, dlookr")

pacman::p_load(tidyverse,
               tidytuesdayR,
               formattable,
               dlookr)
```

### Dataset

```{r}
longbeach <- tt_load("2025-03-04")
longbeach = longbeach$longbeach
```

```{r}
#| label: load-dataset
#| message: false

# Column Names
names(longbeach)

```

This TidyTuesday dataset is a conglomeration of different species retrieved in Long Beach, CA. It contains 22 columns and 29,787 rows that provides potential insights of rescued animals at the city's shelter. The data includes variables such as intake type, outcome type, animal type, age, breed, and geographical jurisdiction. We chose this dataset because it offers an opportunity to explore animal shelters and outcomes. Meaningful and relevant analysis could have direct implications for animal welfare policies and shelter resource management.

### Questions

1.  How do intake conditions, animal types, and ages affect the animal's outcome?

2.  Over the years 2021–2025, how have different species and intake conditions influenced the duration from intake to adoption?

## Analysis plan

For each question, we plan the following:

### 1. How do intake conditions, animal types, and ages affect the animal's outcome?

We will model `outcome_type` as a function of `age_at_intake`, `intake_condition`, and `animal_type` using a linear model (subject to change).

Independent Variables:

-   `age_at_intake` (calculated by `dob` and `intake_date`)

-   `intake_condition`

-   `animal_type`

Dependent Variables:

-   `outcome_type`

#### Data Cleaning

```{r}
longbeach |>
  diagnose() |>
  filter(variables %in% c("dob", "intake_date", "intake_condition", "animal_type", "outcome_type")) |>
  formattable()
```

#### Proposed Visualizations

### 2. Over the years 2021–2025, how have different species and intake conditions influenced the duration from intake to adoption?

We'll mutate() a new variable `days_to_adoption` as the time difference between intake and outcome dates, then analyze its variation by `intake_year` and `outcome_type` across years. Because dogs and cats dominate the data, we collapse all other species into a three-level factor `species_group` (“Dog,” “Cat,” “Other”) for clearer comparisons.

Independent Variables:

-   `animal_type`

-   `intake_condition`

-   `intake_year` (derived from `intake_date`)

-   `outcome_type`

Dependent Variables:

-   `days_to_adoption` (`computed by outcome_date` - `intake_date`)

#### Data Cleaning

```{r}
longbeach |>
  diagnose() |>
  filter(variables %in% c("animal_type", "intake_condition", "intake_date", "outcome_type")) |>
  formattable()
```

```{r}
longbeach_clean <- longbeach |>
  
  # standardize intake_condition
  mutate(
    intake_condition = intake_condition |>
      str_squish()             # collapse multiple spaces
      |> str_replace_all("/", " ")  # replace slashes with spaces
      |> str_to_lower()             # convert to lowercase
  ) |>
  
  # create a simplified species_group
  mutate(
    species_group = case_when(
      animal_type == "dog" ~ "Dog",
      animal_type == "cat" ~ "Cat",
      TRUE ~ "Other"
    ) |> 
      factor(levels = c("Dog", "Cat", "Other"))
  ) |>
  
  # compute age_at_intake and flag unknowns
  mutate(
    age_at_intake = as.numeric(intake_date - dob) / 365,
    age_unknown   = is.na(age_at_intake)
  ) |>
  
  # compute days_to_adoption
  mutate(
    days_to_adoption = as.integer(outcome_date - intake_date)
  )

glimpse(longbeach_clean)
```

#### Proposed Visualizations

Time‐Series Lines: Median `days_to_adoption` by `intake_year` (2021–2025), faceted by `species_group`.

Violin Distribution: Yearly distribution of `days_to_adoption`, split by `intake_condition`

## Weekly Plan of Attack

| Task Name | Status | Due | Priority | Summary |
|---------------|---------------|---------------|---------------|---------------|
| Analysis of dataset and project decision | Complete | 2025-06-13 | High | Team found dataset, selected questions. Discussed approach and division of work |
| Data ingestion & cleaning | Started | 2025-06-13 | High | Load data; inspect structure; derive and clean data |
| Q1 & Q2 plot exploration and decision | Started | 2025-06-13 | High | Decide on plots and narrative |
| Draft narrative & interpretation | Not started | 2025-06-20 | High | Summarize results; write narrative for both questions |
| Finalize report & slides | Not started | 2025-06-30 | High | Refine visuals; polish text; build and rehearse presentation |
