---
title: "Graphically pawsitive vibes"
subtitle: "Proposal"
author: 
  - name: "Isle of Dogs"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project will graphically depict trends, patterns, and analysis of Animal Shelter data derived from the TidyTuesday initiative. The dataset includes information about animals taken in and released from the shelter; there are variables such as intake type, outcome type, animal type, breed, age, and more. The primary goal of this project is to identify key factors that influence an animals outcome."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
#| echo: false

if (!require("pacman")) 
  install.packages("pacman")

pacman::p_load(tidyverse,tidytuesdayR)

longbeach <- tt_load("2025-03-04")
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
#| echo: false

glimpse(longbeach$longbeach)

length(unique(longbeach$longbeach$jurisdiction))
unique(longbeach$longbeach$jurisdiction)

length(unique(longbeach$longbeach$animal_type))
unique(longbeach$longbeach$animal_type)

```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

This TidyTuesday dataset is a conglomeration of different species retrieved in Long Beach, CA. It contains 22 columns and 29,787 rows that provides potential insights of rescued animals at the city's shelter. The data includes variables such as intake type, outcome type, animal type, age, breed, and geographical jurisdiction. We chose this dataset because it offers an opportunity to explore animal shelters and outcomes. Meaningful and relevant analysis could have direct implications for animal welfare policies and shelter resource management.

Make sure to load the data and use inline code for some of this information.

## Questions

1.  How do intake conditions, animal types, and ages affect the animal's outcome?

2.  Over the years 2021–2025, how have different species and intake conditions influenced the duration from intake to adoption?

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
