---
title: "Graphically pawsitive vibes"
subtitle: "Proposal"
author: 
  - name: "Isle of Dogs"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project will graphically depict trends, patterns, and analysis of Animal Shelter data derived from the TidyTuesday initiative. The dataset includes information about animals taken in and released from the shelter; there are variables such as intake type, outcome type, animal type, breed, age, and more. The primary goal of this project is to identify key factors that influence an animals outcome."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
#| echo: false

if (!require("pacman")) 
  install.packages("pacman")

pacman::p_load(tidyverse,tidytuesdayR)

longbeach <- tt_load("2025-03-04")
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
#| echo: false

glimpse(longbeach$longbeach)
summary(longbeach$longbeach)

length(unique(longbeach$longbeach$jurisdiction))
unique(longbeach$longbeach$jurisdiction)

length(unique(longbeach$longbeach$animal_type))
unique(longbeach$longbeach$animal_type)

```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

This TidyTuesday dataset is a conglomeration of different species retrieved in Long Beach, CA. It contains 22 columns and 29,787 rows that provides potential insights of rescued animals at the city's shelter. The data includes variables such as intake type, outcome type, animal type, age, breed, and geographical jurisdiction. We chose this dataset because it offers an opportunity to explore animal shelters and outcomes. Meaningful and relevant analysis could have direct implications for animal welfare policies and shelter resource management.

Make sure to load the data and use inline code for some of this information.

## Questions

1.  How do intake conditions, animal types, and ages affect the animal's outcome?

2.  Over the years 2021–2025, how have different species and intake conditions influenced the duration from intake to adoption?

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

For each question, we plan the following: 

### 1. We will model $y_1$ as a function of $x_1$, $x_2$, and $x_3$ using a linear model (subject to change). 


Variables: 

- **$x_1$ = age** 

- **$x_2$ = intake_type**  

- **$x_3$ = animal_type**

- **$y_1$ = outcome type**

### 2. We'll mutate() a new variable $y_2$ as the time difference between intake and outcome dates, then analyze its variation by $x_4$ and $x_5$ across years.


Variables: 

- **$x_4$ = animal_type**

- **$x_5$ = intake_conditions**

- **$y_2$ = intake_outcome_duration** (requires mutate())

## Plan of Attack 

---

### Week 1: June 8 – June 14  
**Objective: Data Exploration & Cleaning**

- Load and examine the structure of the dataset

- Data Wrangle, filter, and prepare variables

- Create exploratory plots

- Begin cleaning and formatting date columns for mutation

- Start presentation.qmd and write-up in index.qmd (last item)

---

### Week 2: June 15 – June 21  
**Objective: Question 1 & Question 2 deep-dive**

- Complete Index.qmd

- Visualize model predictions by key features and trends over time

- (Q1) Filter and prepare variables: age, intake_type, animal_type, outcome_type

- (Q1) Fit Linear Model

- (Q2) Explore distributions of durations by animal_type and intake_condition

- (Q2) Create `y_2` by mutating: `intake_outcome_duration = outcome_date - intake_date`


---

### Week 3: June 22 – June 25  
**Objective:  Translate Findings into Presentation**

- Refine visualizations and model summaries

- Interpret findings for both questions

- Peer review & revise before submission/presentation (last item)


---

### Week 4: June 26 – June 30  
**Objective: Finalize & Present**

- Organize and clean report

- Record presentation

---
