---
title: "Graphically pawsitive vibes"
subtitle: "Proposal"
author: 
  - name: "Isle of Dogs"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project will graphically depict trends, patterns, and analysis of Animal Shelter data derived from the TidyTuesday initiative. The dataset includes information about animals taken in and released from the shelter; there are variables such as intake type, outcome type, animal type, breed, age, and more. The primary goal of this project is to identify key factors that influence an animals outcome."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
#| echo: false

if (!require("pacman")) 
  install.packages("pacman")

pacman::p_load(tidyverse,tidytuesdayR)

longbeach <- tt_load("2025-03-04")
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

# Column Names
names(longbeach$longbeach)

```

This TidyTuesday dataset is a conglomeration of different species retrieved in Long Beach, CA. It contains 22 columns and 29,787 rows that provides potential insights of rescued animals at the city's shelter. The data includes variables such as intake type, outcome type, animal type, age, breed, and geographical jurisdiction. We chose this dataset because it offers an opportunity to explore animal shelters and outcomes. Meaningful and relevant analysis could have direct implications for animal welfare policies and shelter resource management.


## Questions

1.  How do intake conditions, animal types, and ages affect the animal's outcome?

2.  Over the years 2021–2025, how have different species and intake conditions influenced the duration from intake to adoption?

## Analysis plan

For each question, we plan the following: 

### 1. How do intake conditions, animal types, and ages affect the animal's outcome?

We will model $y_1$ as a function of $x_1$, $x_2$, and $x_3$ using a linear model (subject to change). 

Variables: 

- **$x_1$ = age_at_intake** (calculated by dob and intake_date)

- **$x_2$ = intake_condition**  

- **$x_3$ = animal_type**

- **$y_1$ = outcome type**

### 2. Over the years 2021–2025, how have different species and intake conditions influenced the duration from intake to adoption?

We'll mutate() a new variable $y_2$ as the time difference between intake and outcome dates, then analyze its variation by $x_4$ and $x_5$ across years.


Variables: 

- **$x_4$ = animal_type**

- **$x_5$ = intake_conditions**

- **$x_6$ = intake_year** (derived from intake_date)

- **$x_7$ = outcome_type**

- **$y_2$ = days_to_adoption** (computed by outcome_date - intake_date)

## Plan of Attack 

<table class="table" border="1" cellspacing="0" cellpadding="6">
  <thead>
    <tr>
      <th>Week</th>
      <th>Task Name</th>
      <th>Status</th>
      <th>Due</th>
      <th>Priority</th>
      <th>Summary</th>
    </tr>
  </thead>
  <tbody>
    <!-- Week 1 -->
    <tr>
      <td>Week 1 (June 8–14)</td>
      <td>Load and examine the structure of the dataset</td>
      <td>☐</td>
      <td>June 10</td>
      <td>High</td>
      <td>Initial data understanding and variable types</td>
    </tr>
    <tr>
      <td></td>
      <td>Wrangle and filter variables</td>
      <td>☐</td>
      <td>June 11</td>
      <td>High</td>
      <td>Prepare key variables for analysis</td>
    </tr>
    <tr>
      <td></td>
      <td>Create exploratory plots</td>
      <td>☐</td>
      <td>June 12</td>
      <td>Moderate</td>
      <td>Visual overview of distributions and outliers</td>
    </tr>
    <tr>
      <td></td>
      <td>Format date columns for mutation</td>
      <td>☐</td>
      <td>June 13</td>
      <td>Moderate</td>
      <td>Prepare date variables for duration calculation</td>
    </tr>
    <tr>
      <td></td>
      <td>Complete <code>Index.qmd</code></td>
      <td>☐</td>
      <td>June 14</td>
      <td>High</td>
      <td>Main project landing document</td>
    </tr>
    <tr>
      <td></td>
      <td>Peer review & revise</td>
      <td>☐</td>
      <td>June 14</td>
      <td>Low</td>
      <td>Get feedback and apply revisions</td>
    </tr>

    <!-- Week 2 -->
    <tr>
      <td>Week 2 (June 15–21)</td>
      <td>Visualize model predictions and trends</td>
      <td>☐</td>
      <td>June 16</td>
      <td>Moderate</td>
      <td>Trends across features over time</td>
    </tr>
    <tr>
      <td></td>
      <td>Prepare variables for Q1 (age, intake_type, animal_type, outcome_type)</td>
      <td>☐</td>
      <td>June 17</td>
      <td>High</td>
      <td>Filter necessary data for modeling</td>
    </tr>
    <tr>
      <td></td>
      <td>Fit linear model for Q1</td>
      <td>☐</td>
      <td>June 18</td>
      <td>High</td>
      <td>Analyze how predictors affect outcomes</td>
    </tr>
    <tr>
      <td></td>
      <td>Explore durations by animal_type and intake_condition (Q2)</td>
      <td>☐</td>
      <td>June 19</td>
      <td>Moderate</td>
      <td>Assess time to adoption across groups</td>
    </tr>
    <tr>
      <td></td>
      <td>Create <code>y_2</code> as intake_outcome_duration</td>
      <td>☐</td>
      <td>June 20</td>
      <td>High</td>
      <td>Derived time variable for Q2 analysis</td>
    </tr>
    <tr>
      <td></td>
      <td>Complete <code>Presentation.qmd</code></td>
      <td>☐</td>
      <td>June 21</td>
      <td>High</td>
      <td>Initial draft of final presentation</td>
    </tr>

    <!-- Week 3 -->
    <tr>
      <td>Week 3 (June 22–25)</td>
      <td>Refine visuals, model summaries, and presentation</td>
      <td>☐</td>
      <td>June 23</td>
      <td>High</td>
      <td>Polish charts and summaries for clarity</td>
    </tr>
    <tr>
      <td></td>
      <td>Interpret findings for both questions</td>
      <td>☐</td>
      <td>June 24</td>
      <td>Moderate</td>
      <td>Turn results into clear conclusions</td>
    </tr>

    <!-- Week 4 -->
    <tr>
      <td>Week 4 (June 26–30)</td>
      <td>Organize and clean report</td>
      <td>☐</td>
      <td>June 28</td>
      <td>High</td>
      <td>Prepare final version of report</td>
    </tr>
    <tr>
      <td></td>
      <td>Record presentation</td>
      <td>☐</td>
      <td>June 30</td>
      <td>High</td>
      <td>Deliverable for final submission</td>
    </tr>
  </tbody>
</table>
<!-- Table Produced by ChatGPT-->
